---
title: "New analysis"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message=F)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set(fig.width=9, fig.height=9)
```

<!-- #=============================================================================== -->
#       Load libraries
<!-- #=============================================================================== -->
```{r}
## Load required libraries
library(DESeq2)
library(data.table)
library(tidyverse)
library(vegan)
library(lmPerm)
library(viridis)
library(grid)
library(gridExtra)
library(cowplot)
# library(devtools)
# install_github("eastmallingresearch/Metabarcoding_pipeline/scripts")
library(metafuncs)
```

<!-- #=============================================================================== -->
#       Functions and constants
<!-- #=============================================================================== -->
```{r}
# Run constants
Factor1="Status"
Factor2="Treatment
Design=~Block+Treatment*Status

# colour blind palette
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
    "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

## Custom functions

gfunc <- function(countData,coldata,title) {
#### Rarefaction curve plotter ####  
  
  colData <- colData[names(countData),]
  
  # descending order each sample 
  DT <- data.table(apply(countData,2,sort,decreasing=T))
  
  # get cummulative sum of each sample
  DT <- cumsum(DT)    
  
  # log the count values                            
  DT <- log10(DT)
  
  # set values larger than maximum for each column to NA
  DT <- data.table(apply(DT,2,function(x) {x[(which.max(x)+1):length(x)]<- NA;x}))
  
  # remove rows with all NA
  DT <- DT[rowSums(is.na(DT)) != ncol(DT), ]
  
  # add a count column to the data table
  DT$x <- seq(1,nrow(DT))
  
  # melt the data table for easy plotting 
  MDT <- melt(DT,id.vars="x")
  
  # create an empty ggplot object from the data table
  g <- ggplot(data=MDT,aes(x=x,y=value,colour=variable))
  
  # remove plot background and etc.
  g <- g + theme_classic_thin() %+replace% 
    theme(legend.position="none",axis.title=element_blank())
  
  # plot cumulative reads
  g <- g + geom_line(size=1.5) + scale_colour_viridis(discrete=T)
  
  # add axis lables
  g <- g + ggtitle(title)
  #g <- g + ylab(expression("Log"[10]*" aligned sequenecs"))+xlab("OTU count")
  
  # print the plot
  g
}

```


<!-- #=============================================================================== -->
#       Load data
<!-- #=============================================================================== -->
```{r}
# Load 
ubiome_BAC <- loadData("BAC.otu_table.txt","colData","BAC.sintax.taxa",RHB="BAC")
ubiome_FUN <- loadData("FUN.otu_table.txt","colData","FUN.sintax.taxa",RHB="FUN")

```

<!-- #=============================================================================== -->
#       Create DEseq objects
<!-- #=============================================================================== -->
```{r}

# filtered for minimum of 1000 reads
ubiome_FUN$dds2 <- ubiom_to_des(ubiome_FUN,filter=expression(colSums(countData)>=1000))
ubiome_BAC$dds2 <- ubiom_to_des(ubiome_BAC,filter=expression(colSums(countData)>=1000))

```

<!-- #=============================================================================== -->
#       Sample rarefaction plots
<!-- #=============================================================================== -->        

```{r}
invisible(mapply(assign,names(ubiome_BAC),ubiome_BAC,MoreArgs=list(envir = globalenv())))
g1 <- gfunc(as.data.frame(counts(dds)),as.data.frame(colData(dds)),"Bacteria")
invisible(mapply(assign,names(ubiome_FUN),ubiome_FUN,MoreArgs=list(envir = globalenv())))
g2 <-gfunc(as.data.frame(counts(dds)),as.data.frame(colData(dds)),"Fungi")

grid.arrange(g1,g2,
             left=textGrob(label=expression("Log"[10] * " aligned sequenecs"),rot=90),
             bottom="OTU count",nrow=2)                        
```

<!-- #=============================================================================== -->
#      ***FUNGI***
<!-- #=============================================================================== -->
```{r}
# Fungi
invisible(mapply(assign,names(ubiome_FUN),ubiome_FUN,MoreArgs=list(envir=globalenv())))

```

<!-- #=============================================================================== -->
#       Alpha diversity analysis
<!-- #=============================================================================== -->

## Alpha diversity plot
```{r}

# plot alpha diversity - plot_alpha will convert normalised abundances to integer values
plot_alpha(counts(dds,normalize=T),
           colData(dds),
           design=Factor1,
           colour=Factor2,
           measures=c("Chao1", "Shannon", "Simpson","Observed"),
           type="box") + 
  scale_colour_manual(values=cbPalette)+theme(axis.title.x =  element_blank())
```

## Permutation based anova on diversity index ranks
```{r}
# get the diversity index data
all_alpha_ord <- plot_alpha(counts(dds,normalize=T),colData(dds),design="Status",returnData=T)

# join diversity indices and metadata
all_alpha_ord <- all_alpha_ord[as.data.table(colData(dds),keep.rownames="Samples"),
                               on="Samples"]

```

### Chao1
```{r}
setkey(all_alpha_ord,S.chao1)
all_alpha_ord[,measure:=as.numeric(as.factor(S.chao1))]
summary(aovp(update(design,measure~.),all_alpha_ord,seqs=T))
```

### Shannon
```{r}
setkey(all_alpha_ord,shannon)
all_alpha_ord[,measure:=as.numeric(as.factor(shannon))]
summary(aovp(update(design,measure~.),all_alpha_ord,seqs=T))
```  

### Simpson
```{r}
setkey(all_alpha_ord,simpson)
all_alpha_ord[,measure:=as.numeric(as.factor(shannon))]
summary(aovp(update(design,measure~.),all_alpha_ord,seqs=T))
```

<!-- #=============================================================================== -->
#       Filter data
<!-- #============================================================================ -->

```{r}
dds <- dds[rowSums(counts(dds, normalize=T))>5,]
```
<!-- #=============================================================================== -->
#       Beta diversity PCA/NMDS
<!-- #=============================================================================== -->

## PCA 
```{r}
### PCA ###

# perform PC decomposition of DES object
mypca <- des_to_pca(dds)

# to get pca plot axis into the same scale create a dataframe of PC scores multiplied by their variance
d <-t(data.frame(t(mypca$x)*mypca$percentVar))
```

### Percent variation in first 4 PCs 
```{r}
round(mypca$percentVar[1:4],3)
```

### ANOVA of first 4 PCs
```{r} 
apply(mypca$x[,1:4],2,function(x){
  summary(aov(update(design,x~.),data=cbind(pc,colData(dds))))
})
```

### PCA plot
```{r, fig.width=8,fig.height=5}

plotOrd(d,
        colData(dds),
        design=Factor1,
        shape=Factor2,
        axes=c(2,3),
   #     facet="field_pair", 
        cbPalette=T,
        alpha=0.75,) #+ 
  #geom_line(aes(group=facet),alpha=0.25,linetype=3,colour="#000000") + 
  #theme(text = element_text(size=14))

```

### PCA sum of squares (% var)
```{r}
sum_squares <- apply(mypca$x,2,function(x) 
  summary(aov(update(design,x~.),data=cbind(x,colData(dds))))[[1]][2]
)
sum_squares <- do.call(cbind,sum_squares)
x<-t(apply(sum_squares,2,prop.table))
perVar <- x * mypca$percentVar
#colSums(perVar)
round(colSums(perVar)/sum(colSums(perVar))*100,3)
```

## ADONIS
```{r}
vg <- vegdist(t(counts(dds,normalize=T)),method="bray")
set.seed(sum(utf8ToInt("Xiangming Xu")))
adonis(update(design,vg~.),colData(dds),permutations = 1000)

```

## NMDS ordination
```{r}
set.seed(sum(utf8ToInt("Xiangming Xu")))
ord <- metaMDS(vg,trace=0) 
#sratmax=20000,maxit=20000,try = 177, trymax = 177

nmds <- scores(ord)

g <- plotOrd(nmds,colData(dds),design=Factor1,shape=Factor2,alpha=0.75,cbPalette=T)
g + theme(text=element_text(size=14))
```

### NMDS with phylum or class arrows 
```{r} 
otus <- scores(ord,"species") 

taxmerge <-data.table(inner_join(data.table(OTU=rownames(otus),as.data.frame(otus)),data.table(OTU=rownames(taxData),taxData))) 
taxmerge$phy <- taxaConfVec(taxmerge[,c(-1:-3,-8)],conf=0.9,level=which(colnames(taxmerge[,c(-1:-3,-8)])=="phylum"))
taxmerge$cls <- taxaConfVec(taxmerge[,c(-1:-3,-8)],conf=0.9,level=which(colnames(taxmerge[,c(-1:-3,-8)])=="class")) 

phy <- taxmerge[,lapply(.SD,mean),by=phy,.SDcols=c("NMDS1","NMDS2")]
cls <- taxmerge[,lapply(.SD,mean),by=cls,.SDcols=c("NMDS1","NMDS2")]

g + geom_segment(inherit.aes = F,data=phy,aes(xend=NMDS1,yend=NMDS2,x=0,y=0),size=1.5,arrow=arrow()) + 
 geom_text(inherit.aes = F,data=phy,aes(x=NMDS1,y=(NMDS2+sign(NMDS2)*0.05),label=phy)) 
``` 


<!-- #=============================================================================== -->
#       differential analysis
<!-- #=============================================================================== -->

## DESeq design 
```{r}

# p value for FDR cutoff
alpha <- 0.1

# add design to dds object
design(dds) <- design

# run model
dds <- DESeq(dds,parallel=F)

# build results table
res <- results(dds,alpha=alpha,contrast=c("Status","Diseased","Healthy"))

```

## Result summary

Rank is lowest taxonomic rank with >=0.65 confidence

### All samples
```{r}
summary(res)

# merge DESeq results with taxonomy
res.merge <- as.data.table(res,keep.rownames="OTU")[
  as.data.table(taxData,keep.rownames="OTU"),on="OTU"]

# print sig. results
res.merge[padj<=alpha,.(OTU,rank,
                        baseMean=round(baseMean,2),
                        FC=round(log2FoldChange,2),
                        padj=round(padj,4))][order(FC,decreasing = T),]

```

<!-- ### Write results -->
<!-- ```{r} -->

<!-- fwrite(res.merge,paste(RHB,"_DES.txt"),quote=F,na="",sep="\t") -->
<!-- fwrite(res2.merge,paste(RHB,"_DES_Paired.txt"),quote=F,na="",sep="\t") -->

<!-- ``` -->


#      ***BACTERIA***


<!-- #=============================================================================== -->
```{r}
# Bacteria
invisible(mapply(assign, names(ubiome_BAC), ubiome_BAC, MoreArgs=list(envir = globalenv())))

# add nested pair column
dds$field_pair <- as.factor(paste(dds$Field,dds$Pair,sep="_"))
colData$field_pair <- as.factor(paste(colData$Field,colData$Pair,sep="_"))

## collapse replicates
dds <- collapseReplicates(dds,groupby = paste0(dds$Status,dds$field_pair))
sizeFactors(dds) <- sizeFactors(estimateSizeFactors(dds))

## remove chloroplasts/mitochondria
dds <- dds[!rownames(dds)%in%c(rownames(taxData[taxData$rank=="Streptophyta(g)",]),"OTU3")]

## filter
keep <- c(which(duplicated(colData(dds)$field_pair,fromLast=T)),
          which(duplicated(colData(dds)$field_pair,fromLast=F)))

```

<!-- #=============================================================================== -->
#       Alpha diversity analysis
<!-- #=============================================================================== -->

## Alpha diversity plot
```{r}

# plot alpha diversity - plot_alpha will convert normalised abundances to integer values
plot_alpha(counts(dds,normalize=T),
           colData(dds),
           design=Factor1,
           colour=Factor2,
           measures=c("Chao1", "Shannon", "Simpson","Observed"),
           type="box") + 
  scale_colour_manual(values=cbPalette)+theme(axis.title.x =  element_blank())
```

## Permutation based anova on diversity index ranks
```{r}
# get the diversity index data
all_alpha_ord <- plot_alpha(counts(dds,normalize=T),colData(dds),design="Status",returnData=T)

# join diversity indices and metadata
all_alpha_ord <- all_alpha_ord[as.data.table(colData(dds),keep.rownames="Samples"),
                               on="Samples"]

```

### Chao1
```{r}
setkey(all_alpha_ord,S.chao1)
all_alpha_ord[,measure:=as.numeric(as.factor(S.chao1))]
summary(aovp(update(design,measure~.),all_alpha_ord,seqs=T))
```

### Shannon
```{r}
setkey(all_alpha_ord,shannon)
all_alpha_ord[,measure:=as.numeric(as.factor(shannon))]
summary(aovp(update(design,measure~.),all_alpha_ord,seqs=T))
```  

### Simpson
```{r}
setkey(all_alpha_ord,simpson)
all_alpha_ord[,measure:=as.numeric(as.factor(shannon))]
summary(aovp(update(design,measure~.),all_alpha_ord,seqs=T))
```

<!-- #=============================================================================== -->
#       Filter data
<!-- #============================================================================ -->

```{r}
dds <- dds[rowSums(counts(dds, normalize=T))>5,]
```
<!-- #=============================================================================== -->
#       Beta diversity PCA/NMDS
<!-- #=============================================================================== -->

## PCA 
```{r}
### PCA ###

# perform PC decomposition of DES object
mypca <- des_to_pca(dds)

# to get pca plot axis into the same scale create a dataframe of PC scores multiplied by their variance
d <-t(data.frame(t(mypca$x)*mypca$percentVar))
```

### Percent variation in first 4 PCs 
```{r}
round(mypca$percentVar[1:4],3)
```

### ANOVA of first 4 PCs
```{r} 
apply(mypca$x[,1:4],2,function(x){
  summary(aov(update(design,x~.),data=cbind(pc,colData(dds))))
})
```

### PCA plot
```{r, fig.width=8,fig.height=5}

plotOrd(d,
        colData(dds),
        design=Factor1,
        shape=Factor2,
        axes=c(2,3),
   #     facet="field_pair", 
        cbPalette=T,
        alpha=0.75,) #+ 
  #geom_line(aes(group=facet),alpha=0.25,linetype=3,colour="#000000") + 
  #theme(text = element_text(size=14))

```

### PCA sum of squares (% var)
```{r}
sum_squares <- apply(mypca$x,2,function(x) 
  summary(aov(update(design,x~.),data=cbind(x,colData(dds))))[[1]][2]
)
sum_squares <- do.call(cbind,sum_squares)
x<-t(apply(sum_squares,2,prop.table))
perVar <- x * mypca$percentVar
#colSums(perVar)
round(colSums(perVar)/sum(colSums(perVar))*100,3)
```

## ADONIS
```{r}
vg <- vegdist(t(counts(dds,normalize=T)),method="bray")
set.seed(sum(utf8ToInt("Xiangming Xu")))
adonis(update(design,vg~.),colData(dds),permutations = 1000)

```

## NMDS ordination
```{r}
set.seed(sum(utf8ToInt("Xiangming Xu")))
ord <- metaMDS(vg,trace=0) 
#sratmax=20000,maxit=20000,try = 177, trymax = 177

nmds <- scores(ord)

g <- plotOrd(nmds,colData(dds),design=Factor1,shape=Factor2,alpha=0.75,cbPalette=T)
g + theme(text=element_text(size=14))
```

### NMDS with phylum or class arrows 
```{r} 
otus <- scores(ord,"species") 

taxmerge <-data.table(inner_join(data.table(OTU=rownames(otus),as.data.frame(otus)),data.table(OTU=rownames(taxData),taxData))) 
taxmerge$phy <- taxaConfVec(taxmerge[,c(-1:-3,-8)],conf=0.9,level=which(colnames(taxmerge[,c(-1:-3,-8)])=="phylum"))
taxmerge$cls <- taxaConfVec(taxmerge[,c(-1:-3,-8)],conf=0.9,level=which(colnames(taxmerge[,c(-1:-3,-8)])=="class")) 

phy <- taxmerge[,lapply(.SD,mean),by=phy,.SDcols=c("NMDS1","NMDS2")]
cls <- taxmerge[,lapply(.SD,mean),by=cls,.SDcols=c("NMDS1","NMDS2")]

g + geom_segment(inherit.aes = F,data=phy,aes(xend=NMDS1,yend=NMDS2,x=0,y=0),size=1.5,arrow=arrow()) + 
 geom_text(inherit.aes = F,data=phy,aes(x=NMDS1,y=(NMDS2+sign(NMDS2)*0.05),label=phy)) 
``` 


<!-- #=============================================================================== -->
#       differential analysis
<!-- #=============================================================================== -->

## DESeq design 
```{r}

# p value for FDR cutoff
alpha <- 0.1

# add design to dds object
design(dds) <- design

# run model
dds <- DESeq(dds,parallel=F)

# build results table
res <- results(dds,alpha=alpha,contrast=c("Status","Diseased","Healthy"))

```

## Result summary

Rank is lowest taxonomic rank with >=0.65 confidence

### All samples
```{r}
summary(res)

# merge DESeq results with taxonomy
res.merge <- as.data.table(res,keep.rownames="OTU")[
  as.data.table(taxData,keep.rownames="OTU"),on="OTU"]

# print sig. results
res.merge[padj<=alpha,.(OTU,rank,
                        baseMean=round(baseMean,2),
                        FC=round(log2FoldChange,2),
                        padj=round(padj,4))][order(FC,decreasing = T),]

```

<!-- ### Write results -->
<!-- ```{r} -->

<!-- fwrite(res.merge,paste(RHB,"_DES.txt"),quote=F,na="",sep="\t") -->
<!-- fwrite(res2.merge,paste(RHB,"_DES_Paired.txt"),quote=F,na="",sep="\t") -->

<!-- ``` -->
